Grove: A Convergent Collaborative Structure-Editor Calculus

### 2021-02-16 grove-paper.tex
comment out old equations
add \id macro for grayed UUIDs

\newcommand{\conflict}[2]{\llparenthesis#1|#2\rrparenthesis}

@@

\tikzset{
  %every picture/.style={framed, background rectangle/.style={draw=gray!50}},
}
\tikzset{edge style/.style={
  draw,
  %color=gray,
  font={\small\ttfamily},
  /tikz/every edge quotes/.style={
    %draw=gray!20,
    anchor=west,
    swap/.append code={
      \ifpgfarrowswap
        \pgfkeysalso{anchor=west}
      \else
        \pgfkeysalso{anchor=east}
      \fi}},
}}
\tikzset{graphs/graph style/.style={
  tree layout,
  level distance=0.5cm,
  level sep=0.5cm,
  sibling distance=0.5cm,
  sibling sep=0.1cm,
  part distance=0.1cm,
  part sep=0.1cm,
  component distance=0.1cm,
  component sep=0.1cm,
  nodes={
    draw,
    %color=gray,
    inner sep=2pt,
    rounded corners=1mm},
  edges={edge style},
}}
\tikzset{graphs/root style/.style={
 %draw=none,
 as={\textbullet$_{0}$}
}}
\tikzset{alice/.style={
  color=red!80!black,
  font={\bfseries\small},
  thick,
}}
\tikzset{bob/.style={
  color=green!60!black,
  font={\bfseries\small},
  thick,
}}
\tikzset{merge/.style={
  color=blue,
  font={\bfseries\small},
  thick,
}}
\tikzset{alice edge/.style={alice, edge style, font={\bfseries\small}}}
\tikzset{alice node/.style={alice}}
\tikzset{alice step/.style={alice}}
\tikzset{bob edge/.style={bob, edge style}, font={\bfseries\small}}
\tikzset{bob node/.style={bob}}
\tikzset{bob step/.style={bob}}
\tikzset{merge edge/.style={merge, edge style}, font={\bfseries\small}}
\tikzset{merge node/.style={merge}}
\tikzset{merge step/.style={merge,decorate,decoration={coil,amplitude=1.0pt,segment length=7.0pt,aspect=0}}}
\tikzset{star/.style={edge node={node[inner sep=0pt,at end,sloped] {\textbf{\huge${}^{\ast}$}}}}}

@@

% Calling \newvertex{Foo}{bar} defines
%   \vidFoo to be a new id number, and
%   \vFoo to be \texttt{bar}\ensuremath{_{\vidFoo}}
\newcounter{NodeVertexCounter}
\newcommand{\newvertex}[2]{%
\ifodd\theNodeVertexCounter
  \addtocounter{NodeVertexCounter}{1}%
\else
  \addtocounter{NodeVertexCounter}{2}%
\fi
\expandafter\newcommand\csname vid#1\endcsname{}% This is just to check if this is a redefinition
\expandafter\global\expandafter\edef\csname vid#1\endcsname{\theNodeVertexCounter}%
\expandafter\newcommand\csname v#1\endcsname{}% This is just to check if this is a redefinition
\expandafter\gdef\csname v#1\endcsname{\texttt{#2}\ensuremath{_{\csname vid#1\endcsname}}}%
}
\newcommand{\newedge}[2]{%
\ifodd\theNodeVertexCounter
  \addtocounter{NodeVertexCounter}{2}%
\else
  \addtocounter{NodeVertexCounter}{1}%
\fi
\expandafter\newcommand\csname eid#1\endcsname{}% This is just to check if this is a redefinition
\expandafter\global\expandafter\edef\csname eid#1\endcsname{\theNodeVertexCounter}%
\expandafter\newcommand\csname e#1\endcsname{}% This is just to check if this is a redefinition
\expandafter\gdef\csname e#1\endcsname{\texttt{#2}\ensuremath{_{\csname eid#1\endcsname}}}%
}
\setcounter{NodeVertexCounter}{-1}
\newvertex{Root}{Root}

@@

\newcommand{\Edge}[1]{Edge~#1}
\newcommand{\Vertex}[1]{Vertex~#1}
\newcommand{\figureCode}[1]{\textbf{\texttt{#1}}\vskip1em}
\newcommand{\multiVertex}[1]{\textcolor{red}{\ensuremath{\mathbf{\curlyveedownarrow_{#1}}}}}
\newcommand{\cycleVertex}[1]{\textcolor{red}{\ensuremath{\mathbf{\rcirclearrowleft_{#1}}}}}
\newcommand{\orphanVertex}[1]{\textcolor{red}{\ensuremath{\mathbf{\ndownarrow_{#1}}}}}
\newcommand{\parens}[1]{\textcolor{gray}{(}#1\textcolor{gray}{)}}
\newcommand{\otherVertexVskip}{\vskip0.5em}

@@

\begin{figure}
  \[
    \arraycolsep=0pt
    \begin{array}{lrlll}
      q    & {}\in Pat & {}::={} & x^u \mid \hole \mid \conflictHole{q,\ldots,q} \mid \multiVertex{u} \mid \cycleVertex{u} \\
      \tau & {}\in Typ & {}::={} & \tau \rightarrow^u \tau \mid Num^u \mid \hole \mid \conflictHole{\tau,\ldots,\tau} \mid \multiVertex{u} \mid \cycleVertex{u} \\
      e    & {}\in Exp & {}::={} & x^u \mid \lambda^u q \mathord{:}\tau.e \mid (e~e)^u \mid n^u \\
           &           &         & ~\mid e~\texttt{+}^u~e \mid e~\texttt{*}^u~e \mid \hole \mid \conflictHole{e,\ldots,e} \mid \multiVertex{u} \mid \cycleVertex{u} \\
    \end{array}
  \]
  \caption{Base syntax, as a grammar, for the lambda calculus that we are considering.}
  \label{fig:Base syntax}
\end{figure}

@@

A grove is a 4-tuple of the following form:
\[
  (e_r, MP, NP, W)
\]
where
\begin{align*}
  e_r &= \expr(v_0) \\
  MP &= \{\expr(v) : |\parents(v)| > 1\} \\
  NP &= \{\expr(v) : |\parents(v)| = 0 \land v \ne v_0\} \\
  W  &= \{\expr(v) : |\parents(v)| = 1 \wedge v = \min(\ancestors(v))\}
\end{align*}
For example, the grove corresponding to the graph in \autoref{fig:Decomposition example} is
\begin{align*}
  e_r &= \hole~\texttt{*}^{40}~\multiVertex{42}~\texttt{+}^{38}~\multiVertex{42} \\
  MP &= \{x^{42}\} \\
  NP &= \{z^{52}\} \\
  W &= \{\hole~\texttt{+}^{46}~\cycleVertex{46}~\texttt{*}^{48}~y^{50}\} \\
\end{align*}

Let us now examine each of the four components in more detail.
The first component $e_r$ is the expression derived from the root vertex $v_0$.
\autoref{fig:Decomposition expr} defines $\expr(v)$ wich determines the expression derived from $v$.
There is one rule for each constructor form.
For example, the expression derived from $v_{38} = (38, \ExpPlus)$ is $\expr'(v_{38}, \PlusLeft) \text{+}^{38} \expr'(v_{38}, \PlusRight)$.
In each child position, $\expr'(v, p)$ determines the sub-expression derived from the children of $v$ at position $p$.
An empty hole $\hole$ appears if there are no children at the given position.
A conflict hole $\conflictHole{e_1,\ldots,e_n}$ appears if there are multiple children.
If there is one child, the expression is determined recursively unless that child appears in $MP$ or $W$, discussed below, in which case we leave a reference, $\multiVertex{u}$ or $\cycleVertex{u}$ respectively.
For the $\ExpLam$ case, we define $\patt(q)$ and $\type(\tau)$ similarly.

The second component $MP$ is the set of expressions derived from a vertex with multiple parents.
Wherever an edge to such a vertex appears, we place a reference to it $\multiVertex{u}$.
In the example, $x^{48}$ is referenced twice in $e_r$.

The third component $NP$ is the set of expressions rooted at a vertex with no parents, other than the root vertex.
In the example, $z^{58}$ has no parents (because it was deleted), so it appears only in $NP$.

@@

\begin{figure}
  \begin{displaymath}
    \begin{array}{l}
      \expr(v=(u, Root)) = expr'(v, Root) \\
      \expr(v=(u, \ExpVar(x))) = x^u \\
      \expr(v=(u, \ExpLam)) = \lambda^u \patt'(v, \LamParam) : \type'(v, \LamType) . \expr'(v, \LamBody) \\
      \expr(v=(u, \ExpApp)) = (\expr'(v, \AppFun)~\expr'(v, \AppArg))^u \\
      \expr(v=(u, \ExpNum(n))) = n^u \\
      \expr(v=(u, \ExpPlus)) = \expr'(v, \PlusLeft)~\texttt{+}^u~\expr'(v, \PlusRight) \\
      \expr(v=(u, \ExpTimes)) = \expr'(v, \TimesLeft)~\texttt{*}^u~\expr'(v, \TimesRight) \\
      \patt(v=(u, \PatVar(x))) = x^u \\
      \type(\tau=(u, \TypArrow)) = \type'(\tau, \ArrowArg) \rightarrow \type'(\tau, \ArrowResult) \\
      \type(\tau=(u, \TypNum) = Num^u \\
    \end{array}
  \end{displaymath}

@@

\[
  \begin{array}{l}
    \textrm{User Actions}: \grave{e} \in Act ::= e \mid \#u \mid \conflict{\grave{e}}{\grave{e}}                                                       \\
    \s : \G \to Act \times \V^{\ast{}} \times \V^{\ast{}} \times \V^{\ast{}}                                                                                          \\
    \s(G) = (\grave{e}, \MP(G), \CC(G), \D(G))                                                                                                         \\
    \MP(G) = \left\{ v \in \vertexes(G) \mid \exists e_1,e_2 \in \liveEdges(G), e_1= \right.                                                           \\
    \qquad \left. (v,p_1,v_1), e_2=(v,p_2,v_2), e_1 \ne e_2 \right\}                                                                                   \\
    \CC(G) = \ldots                                                                                                                                    \\
    \D(G) = \orphans(G) \setminus \left\{ v_0 \right\}                                                                                                 \\
    \orphans(G) = \left\{ v \in \vertexes(G) \mid \forall e=(v^\prime,p,v^{\prime\prime}) \in \right.                                                  \\
    \qquad \left. \liveEdges(G), v^{\prime\prime} \ne v \right\}                                                                                       \\
    \vertexes(G) = \left\{ v \mid \exists e = (v^\prime, p, v^{\prime\prime}) \in \edges(G), v \in \left\{ v^\prime, v^{\prime\prime}\right\} \right\} \\
    \edges(G) = \left\{ e \mid G(e) \ne \bot \right\}                                                                                                  \\
    \liveEdges(G) = \left\{ e \mid G(e) = + \right\}                                                                                                   \\
  \end{array}
\]

>>>>

\newcommand{\id}[1]{\textcolor{gray}{\ensuremath{#1}}}

@@

\tikzset{
  %every picture/.style={framed, background rectangle/.style={draw=gray!50}},
}
\tikzset{edge style/.style={
  draw,
  %color=gray,
  font={\small\ttfamily},
  /tikz/every edge quotes/.style={
    %draw=gray!20,
    anchor=west,
    swap/.append code={
      \ifpgfarrowswap
        \pgfkeysalso{anchor=west}
      \else
        \pgfkeysalso{anchor=east}
      \fi}},
}}
\tikzset{graphs/graph style/.style={
  tree layout,
  level distance=0.5cm,
  level sep=0.5cm,
  sibling distance=0.5cm,
  sibling sep=0.1cm,
  part distance=0.1cm,
  part sep=0.1cm,
  component distance=0.1cm,
  component sep=0.1cm,
  nodes={
    draw,
    %color=gray,
    inner sep=2pt,
    rounded corners=1mm},
  edges={edge style},
}}
\tikzset{graphs/root style/.style={
 %draw=none,
 as={\textbullet$_{\id{0}}$}
}}
\tikzset{alice/.style={
  color=red!80!black,
  font={\bfseries\small},
  thick,
}}
\tikzset{bob/.style={
  color=green!60!black,
  font={\bfseries\small},
  thick,
}}
\tikzset{merge/.style={
  color=blue,
  font={\bfseries\small},
  thick,
}}
\tikzset{alice edge/.style={alice, edge style, font={\bfseries\small}}}
\tikzset{alice node/.style={alice}}
\tikzset{alice step/.style={alice}}
\tikzset{bob edge/.style={bob, edge style}, font={\bfseries\small}}
\tikzset{bob node/.style={bob}}
\tikzset{bob step/.style={bob}}
\tikzset{merge edge/.style={merge, edge style}, font={\bfseries\small}}
\tikzset{merge node/.style={merge}}
\tikzset{merge step/.style={merge,decorate,decoration={coil,amplitude=1.0pt,segment length=7.0pt,aspect=0}}}
\tikzset{star/.style={edge node={node[inner sep=0pt,at end,sloped] {\textbf{\huge${}^{\ast}$}}}}}

@@

% Calling \newvertex{Foo}{bar} defines
%   \vidFoo to be a new id number, and
%   \vFoo to be \texttt{bar}\ensuremath{_{\vidFoo}}
\newcounter{NodeVertexCounter}
\newcommand{\newvertex}[2]{%
\ifodd\theNodeVertexCounter
  \addtocounter{NodeVertexCounter}{1}%
\else
  \addtocounter{NodeVertexCounter}{2}%
\fi
\expandafter\newcommand\csname vid#1\endcsname{}% This is just to check if this is a redefinition
\expandafter\global\expandafter\edef\csname vid#1\endcsname{\theNodeVertexCounter}%
\expandafter\newcommand\csname v#1\endcsname{}% This is just to check if this is a redefinition
\expandafter\gdef\csname v#1\endcsname{\texttt{#2}\ensuremath{_{\id{\csname vid#1\endcsname}}}}%
}
\newcommand{\newedge}[2]{%
\ifodd\theNodeVertexCounter
  \addtocounter{NodeVertexCounter}{2}%
\else
  \addtocounter{NodeVertexCounter}{1}%
\fi
\expandafter\newcommand\csname eid#1\endcsname{}% This is just to check if this is a redefinition
\expandafter\global\expandafter\edef\csname eid#1\endcsname{\theNodeVertexCounter}%
\expandafter\newcommand\csname e#1\endcsname{}% This is just to check if this is a redefinition
\expandafter\gdef\csname e#1\endcsname{\texttt{#2}\ensuremath{_{\id{\csname eid#1\endcsname}}}}%
}
\setcounter{NodeVertexCounter}{-1}
\newvertex{Root}{Root}

@@

\newcommand{\Edge}[1]{Edge~#1}
\newcommand{\Vertex}[1]{Vertex~#1}
\newcommand{\figureCode}[1]{\textbf{\texttt{#1}}\vskip1em}
\newcommand{\multiVertex}[1]{\textcolor{red}{\ensuremath{\curlyveedownarrow_{\id{#1}}}}}
\newcommand{\cycleVertex}[1]{\textcolor{red}{\ensuremath{\rcirclearrowleft_{\id{#1}}}}}
\newcommand{\orphanVertex}[1]{\textcolor{red}{\ensuremath{\mathbf{\ndownarrow_{#1}}}}}
\newcommand{\parens}[1]{\textcolor{gray}{(}#1\textcolor{gray}{)}}
\newcommand{\otherVertexVskip}{\vskip0.5em}

@@

\begin{figure}
  \[
    \arraycolsep=0pt
    \begin{array}{lrlll}
      q    & {}\in Pat & {}::={} & x^{\id{u}} \mid \hole \mid \conflictHole{q,\ldots,q} \mid \multiVertex{u} \mid \cycleVertex{u} \\
      \tau & {}\in Typ & {}::={} & \tau \rightarrow^{\id{u}} \tau \mid Num^{\id{u}} \mid \hole \mid \conflictHole{\tau,\ldots,\tau} \mid \multiVertex{u} \mid \cycleVertex{u} \\
      e    & {}\in Exp & {}::={} & x^{\id{u}} \mid \lambda^{\id{u}} q \mathord{:}\tau.e \mid (e~e)^{\id{u}} \mid n^{\id{u}} \\
           &           &         & ~\mid e~\texttt{+}^{\id{u}}~e \mid e~\texttt{*}^{\id{u}}~e \mid \hole \mid \conflictHole{e,\ldots,e} \mid \multiVertex{u} \mid \cycleVertex{u} \\
    \end{array}
  \]
  \caption{Base syntax, as a grammar, for the lambda calculus that we are considering.}
  \label{fig:Base syntax}
\end{figure}

@@

A grove is a 4-tuple of the following form:
\[
  (e_r, MP, NP, W)
\]
where
\begin{align*}
  e_r &= \expr(v_0) \\
  MP &= \{\expr(v) : |\parents(v)| > 1\} \\
  NP &= \{\expr(v) : |\parents(v)| = 0 \land v \ne v_0\} \\
  W  &= \{\expr(v) : |\parents(v)| = 1 \wedge v = \min(\ancestors(v))\}
\end{align*}
For example, the grove corresponding to the graph in \autoref{fig:Decomposition example} is
\begin{align*}
  e_r &= \hole~\texttt{*}^{\id{40}}~\multiVertex{42}~\texttt{+}^{\id{38}}~\multiVertex{42} \\
  MP &= \{x^{\id{42}}\} \\
  NP &= \{z^{\id{52}}\} \\
  W &= \{\hole~\texttt{+}^{\id{46}}~\cycleVertex{46}~\texttt{*}^{\id{48}}~y^{\id{50}}\} \\
\end{align*}

Let us now examine each of the four components in more detail.
The first component $e_r$ is the expression derived from the root vertex $v_0$.
\autoref{fig:Decomposition expr} defines $\expr(v)$ wich determines the expression derived from $v$.
There is one rule for each constructor form.
For example, the expression derived from $v_{\id{u}} = (u, \ExpPlus)$ is $\expr'(v_{\id{u}}, \PlusLeft) \text{+}^{\id{u}} \expr'(v_{\id{u}}, \PlusRight)$.
In each child position, $\expr'(v, p)$ determines the sub-expression derived from the children of $v$ at position $p$.
An empty hole $\hole$ appears if there are no children at the given position.
A conflict hole $\conflictHole{e_1,\ldots,e_n}$ appears if there are multiple children.
If there is one child, the expression is determined recursively unless that child appears in $MP$ or $W$, discussed below, in which case we leave a reference, $\multiVertex{u}$ or $\cycleVertex{u}$ respectively.
For the $\ExpLam$ case, we define $\patt(q)$ and $\type(\tau)$ similarly.

The second component $MP$ is the set of expressions derived from a vertex with multiple parents.
Wherever an edge to such a vertex appears, we place a reference to it $\multiVertex{u}$.
In the example, $x_{\id{42}}$ is referenced twice in $e_r$.

The third component $NP$ is the set of expressions rooted at a vertex with no parents, other than the root vertex.
In the example, $z^{\id{58}}$ has no parents (because it was deleted), so it appears only in $NP$.

@@

\begin{figure}
  \begin{displaymath}
    \begin{array}{l}
      \expr(v=(u, Root)) = expr'(v, Root) \\
      \expr(v=(u, \ExpVar(x))) = x^{\id{u}} \\
      \expr(v=(u, \ExpLam)) = \lambda^{\id{u}} \patt'(v, \LamParam) : \type'(v, \LamType) . \expr'(v, \LamBody) \\
      \expr(v=(u, \ExpApp)) = (\expr'(v, \AppFun)~\expr'(v, \AppArg))^{\id{u}} \\
      \expr(v=(u, \ExpNum(n))) = n^{\id{u}} \\
      \expr(v=(u, \ExpPlus)) = \expr'(v, \PlusLeft)~\texttt{+}^{\id{u}}~\expr'(v, \PlusRight) \\
      \expr(v=(u, \ExpTimes)) = \expr'(v, \TimesLeft)~\texttt{*}^{\id{u}}~\expr'(v, \TimesRight) \\
      \patt(v=(u, \PatVar(x))) = x^{\id{u}} \\
      \type(\tau=(u, \TypArrow)) = \type'(\tau, \ArrowArg) \rightarrow^{\id{u}} \type'(\tau, \ArrowResult) \\
      \type(\tau=(u, \TypNum) = Num^{\id{u}} \\
    \end{array}
  \end{displaymath}

@@

% \[
%   \begin{array}{l}
%     \textrm{User Actions}: \grave{e} \in Act ::= e \mid \#u \mid \conflict{\grave{e}}{\grave{e}}                                                       \\
%     \s : \G \to Act \times \V^{\ast{}} \times \V^{\ast{}} \times \V^{\ast{}}                                                                                          \\
%     \s(G) = (\grave{e}, \MP(G), \CC(G), \D(G))                                                                                                         \\
%     \MP(G) = \left\{ v \in \vertexes(G) \mid \exists e_1,e_2 \in \liveEdges(G), e_1= \right.                                                           \\
%     \qquad \left. (v,p_1,v_1), e_2=(v,p_2,v_2), e_1 \ne e_2 \right\}                                                                                   \\
%     \CC(G) = \ldots                                                                                                                                    \\
%     \D(G) = \orphans(G) \setminus \left\{ v_0 \right\}                                                                                                 \\
%     \orphans(G) = \left\{ v \in \vertexes(G) \mid \forall e=(v^\prime,p,v^{\prime\prime}) \in \right.                                                  \\
%     \qquad \left. \liveEdges(G), v^{\prime\prime} \ne v \right\}                                                                                       \\
%     \vertexes(G) = \left\{ v \mid \exists e = (v^\prime, p, v^{\prime\prime}) \in \edges(G), v \in \left\{ v^\prime, v^{\prime\prime}\right\} \right\} \\
%     \edges(G) = \left\{ e \mid G(e) \ne \bot \right\}                                                                                                  \\
%     \liveEdges(G) = \left\{ e \mid G(e) = + \right\}                                                                                                   \\
%   \end{array}
% \]
